A concise manual to\\\textsc{DEVS Ex Machina}

### 2015-04-24 manual/tex/advanced.tex

\section{Time}
\paragraph{Explain default time implementation}
\paragraph{Point to where the typedef can be changed}
\paragraph{Explain how users can create their own whacky implementation of the Time class for if the current version does not suit their needs}

\section{Tracing}
\paragraph{Briefly explain where the user can change the tracerset typedef}
\paragraph{Explain individual tracers}
\paragraph{Explain how the user can implement their own Tracer or OutputPolicy}

>>>>

\lstset{
    frameround=fttt,
    language=C++,
    breaklines=true,
}

\section{Time}
Time is a very important aspect of the simulator. Our implementation allows users to easily change the representation of time in order to fit their needs.
\paragraph{Custom time object}
You can change how time is represented by changing the \lstinline{t_timestamp} type definition in the \path{src/forwarddeclare/timestamp.h} header. Please take care that this type implements at least the interface provided by the default \lstinline{Time} class. For more information, take a look at the accompanying documentation.

\section{Tracing}
The tracers generate output during the simulation.
\label{sec:Tracing}
\paragraph{The Tracing interface}
The interface consists of only a handful methods.
\paragraph{Custom set of tracers}
If the default set of tracers is not satisfactory, then you can select your own set of tracers. The header \path{tracers.h} in the folder \path{src/forwarddeclare} contains the typedef for the default implementation.
\begin{lstlisting}
typedef Tracers<VerboseTracer<CoutWriter>> t_tracerset;
\end{lstlisting}
You can select as many tracers as you want by providing more template arguments. The following example sets the simulator to use two tracers. The first tracer will generate the default verbose output and write it to a file. The second tracer is a custom tracer.
\begin{lstlisting}
typedef Tracers<VerboseTracer<FileWriter>,
                MyTracer> t_tracerset;
\end{lstlisting}
Note that some tracers might need special initialization prior to use, such as the tracers provided by us that make use of the \lstinline{FileWriter} policy.

\paragraph{Provided Tracers}
All provided tracers make use of the policy pattern to separate the output generation from sending that output to an output sink. Currently, there is a policy for sending the output directly to standard out (\lstinline{CoutWriter}) or writing it to a file(\lstinline{FileWriter}). Both policies can be found in the \path{src/tracers/policies.h} header. Note that in the case of \lstinline{FileWriter}, the tracer must be explicitly initialized before it can be used. For more implementations, refer to the provided documentation.\\
Currently, the simulator provides only one tracer class: the \lstinline{VerboseTracer}. This tracer will generate verbose output that is comparable to the verbose output of the Python PDEVS implementation by Jentl Van Tendeloo.
\paragraph{Implementing a custom tracer}
\subparagraph{The tracing interface}
The custom tracer must at least implement the following interface:
\begin{lstlisting}
interface Tracer: public TracerBase
{
public:
  /**
   * @brief Traces state initialization of a model
   * @param model The model that is initialized
   * @param time The simulation time of initialization.
   */
  void tracesInit(const t_atomicmodelptr& adevs,
                  t_timestamp time);

  /**
   * @brief Traces internal state transition
   * @param adevs The atomic model that just
   *              performed an internal transition
   * @precondition The model pointer is not a nullptr
   */
  void tracesInternal(const t_atomicmodelptr& adevs,
                      std::size_t coreid);

  /**
   * @brief Traces external state transition
   * @param model The model that just went
   *              through an external transition
   */
  void tracesExternal(const t_atomicmodelptr& adevs,
                      std::size_t coreid);

  /**
   * @brief Traces confluent state transition
   * A confluent state transition is
   * a simultaneous internal and external transition.
   * @param model The model that just went
   *              through a confluent transition
   */
  void tracesConfluent(const t_atomicmodelptr& adevs,
                       std::size_t coreid);
};
\end{lstlisting}
Even though it is not explicitly enforced, it is highly recommended to derive your custom implementation from the \lstinline{TracerBase} class and to allow the use of output policies.\\
Note that the tracer must not print its output upon generation. To make sure that the output is still generated in the correct order, the tracer must register a \lstinline{TraceMessage}.
\subparagraph{Registering a trace message}
The \lstinline{TraceMessage} class is defined in the \path{src/tracers/tracemessage.h} header file. These messages allow the simulator to schedule output generation and purge unwanted messages upon rollback. They contain the following elements:
\begin{itemize}
	\item[timestamp] The time upon which this message is created. This time can be retrieved either from the model that is being traced or as a separate parameter.
	\item[tracerID] Each tracer should have a unique ID. If multiple tracers are used, this ID makes sure that the tracers cannot interfere with eachother. If your tracer derives from \lstinline{TracerBase}, then this ID is automatically assigned to your tracer. It can be accessed via the \lstinline{getID()} member function.
	\item[execute function] This function is executed when the trace output can be safely sent to an output sink. It is recommended to use \lstinline{std::bind<...>()}. The function must be of type \lstinline{void()}.
	\item[coreID] Each simulation core has an ID. During the simulation, it can happen that some trace messages must be removed based on which core initiated the trace.
	\item[delete function] This function is called when the \lstinline{TraceMessage} is destructed. If the tracer must allocate memory on the heap, this function can be used to free that memory.\\*
		Just like the \emph{execute function}, this function must be of type \lstinline{void()}.
\end{itemize}
When the message is created, it can be scheduled with the free function \lstinline{void scheduleMessage(t_tracemessageptr message)}.\\
Both the \lstinline{TraceMessage} class and the \lstinline{scheduleMessage} function are defined in the \path{src/tracers/tracemessage.h} header. For a more in-depth example, you can always take a look at the implementation of the \lstinline{VerboseTracer} implementation.

### 2015-04-24 manual/tex/models.tex

\section{States}
You can can make your own states with almost no limitations. The only requirements for a user's own states are that they should be objects from a type that implements \textsl{n\_model::State}. This class has a few functions that have to be implemented:
\begin{itemize}
	\item \textsl{std::string toString()}
	\item \textsl{std::string toXML()}
	\item \textsl{std::string toCell()}
\end{itemize}
While members in sub-classes can be added without any problems, by default there is a public member \textsl{std::string m\_state} implemented in \textsl{State} that can be used to represent the user's state.

\section{Atomic models}
Atomic models can be made by implementing a sub-class of \\ \textsl{n\_model::AtomicModel}. The most basic functions to be implemented are:
\begin{enumerate}
	\item \textsl{void AtomicModel::extTransition(const std::vector{\textless}n\_network::t\_msgptr{\textgreater} \& inputs)}\\
		Performs the model's external transition. The input argument is a vector with all messages at a certain point in time.
	\item \textsl{void AtomicModel::intTransition()}\\
		Performs the model's internal transition. 
	\item \textsl{t\_timestamp AtomicModel::timeAdvance()}\\
		Should return the model's current time advance value; the amount of time after which the following internal transition occurs. This function should not alter the model in any way.
	\item \textsl{std::vector{\textless}n\_network::t\_msgptr{\textgreater} AtomicModel::output()  const}\\
		Returns the model's output; all messages that are to be send should be returned via this function. This function should not alter the model in any way.
\end{enumerate}

@@

When \textbf{sending messages} (\textbf{output function}), there are a few rules you should take into account. The message you want to send can only be a string. This string should be given to the port you want to send it through. The port will then create the actual messages (that are send over the models) and return them to you in a vector. All these messages (from possibly different ports) should be joined in 1 vector and be returned in the output function. The following functions will allow you to do so:
\begin{itemize}
	\item \textsl{t\_portptr Model::getPort(std::string name) const}
	\item \textsl{std::vector{\textless}n\_network::t\_msgptr{\textgreater} Port:: \\createMessages(std::string message)}
\end{itemize} 
Something like this would work: this-{\textgreater}getPort("OUT")-{\textgreater} \\createMessages(message); \\

@@

\textbf{Connecting these ports} can be done with the function \textsl{void CoupledModel::connectPorts(const t\_portptr\& p1, const t\_portptr\& p2, t\_zfunc zFunction);}. This function connects port \textsl{p1} to port \textsl{p2}. An optional Z function can be added that will modify the messages during transit from one port to another.
It is possible to connect ports from a coupled model to one of it's direct atomic (or coupled) model children.

>>>>

\section{States}
You can can make your own states with almost no limitations. The only requirements for a user's own states are that they should be objects from a type that implements \textsl{n\_model::State}. This class has a few functions that have to be implemented:
\begin{itemize}
	\item \textsl{std::string toString()}
	\item \textsl{std::string toXML()}
	\item \textsl{std::string toCell()}
\end{itemize}
If the user does not override the above functions, this will result in a logged error. Serialization prohibits abstract functions, which enforces this workaround.\\
While members in sub-classes can be added without any problems, by default there is a public member \textsl{std::string m\_state} implemented in \textsl{State} that can be used to represent the user's state.
\\ For serialization you have to declare (after class definition) 
\begin{verbatim}
CEREAL_REGISTER_TYPE(MyState);
\end{verbatim}

\section{Atomic models}
Atomic models can be made by implementing a sub-class of \\ \textsl{n\_model::AtomicModel}. The most basic functions to be implemented are:
\begin{enumerate}
	\item \textsl{void AtomicModel::extTransition(const std::vector{\textless}n\_network::t\_msgptr{\textgreater} \& inputs)}\\
		Performs the model's external transition. The input argument is a vector with all messages at a certain point in time.
	\item \textsl{void AtomicModel::intTransition()}\\
		Performs the model's internal transition. 
	\item \textsl{t\_timestamp AtomicModel::timeAdvance() const}\\
		Should return the model's current time advance value; the amount of time after which the following internal transition occurs. This function should not alter the model in any way.
	\item \textsl{std::vector{\textless}n\_network::t\_msgptr{\textgreater} AtomicModel::output()  const}\\
		Returns the model's output; all messages that are to be send should be returned via this function. This function should not alter the model in any way.
\end{enumerate}
If these are not overridden, this will result in a logged assertion error.\\

@@

When \textbf{sending messages} (\textbf{output function}), there are a few rules you should take into account. The message you want to send can only be a string. This string should be given to the port you want to send it through. The port will then create the actual messages (that are send over the models) and return them to you in a vector. All these messages (from possibly different ports) should be joined in 1 vector and be returned in the output function. The following functions will allow you to do so:
\begin{itemize}
	\item \textsl{t\_portptr Model::getPort(std::string name) const}
	\item \textsl{std::vector{\textless}n\_network::t\_msgptr{\textgreater} Port:: \\createMessages(std::string message)}
\end{itemize} 
Something like this would work: this-{\textgreater}getPort("OUT")-{\textgreater} \\createMessages("my\_message"); \\

It's important that a model name is unique over the whole simulation, and portname should be unique per model.\\

For serialization you have to declare (after class definition)
\begin{verbatim}
CEREAL_REGISTER_TYPE(MyModel);
\end{verbatim}

@@

\textbf{Connecting these ports} can be done with the function\\ \textsl{void CoupledModel::connectPorts(const t\_portptr\& p1, const t\_portptr\& p2, t\_zfunc zFunction);}.\\ This function connects port \textsl{p1} to port \textsl{p2}. An optional Z function can be added that will modify the messages during transit from one port to another.
It is possible to connect ports from a coupled model to one of it's direct atomic (or coupled) model children.

### 2015-04-24 manual/tex/reqs.tex

\section{System requirements}
\subsubsection{Compiler}
The project requires a recent version of a standards-compliant C++11 compiler and runtime. The following are compilers/runtimes used to develop the simulator (and therefore guaranteed to work)
\begin{itemize}
  \item G++ 4.8.2 + libstdc++
  \item G++ 4.9.2 + libstdc++
  \item Clang 3.4.0 + libstdc++
  \item Clang 3.5.0 + libstdc++
  \item G++ 4.9.2 on Cygwin $>=$ 1.7
\end{itemize}
Especially the C++11 threading support is essential.
\subsubsection{Dependencies}
The build system requires CMake $>=$ 2.8, the presence of standard build tools is assumed (make etc.).
The following are the functional dependencies:
\begin{itemize}
  \item Cereal (source included)
  \item GTest $>=$ 1.6
  \item Boost heap $>=$ 1.55  
\end{itemize}
Boost can be installed on *nix with the following command:\\
Debian based:\\
\lstinline!$sudo apt-get install libboost-all-dev!\\
Red-hat based:\\
\lstinline!$sudo yum install boost-devel!\\
\subsubsection{Preparing the build}
We use thread-sanitizer to verify thread-safety at runtime instrumented by the compiler (g++ and clang). This however requires building the project (and by extension any linked libraries) with position independent flags. For the simulator this is resolved by CMake, however for gtest you'll need to manually build GTest with this extra flag:\\
In file internal\textunderscore utils.cmake change compile options to append the compile flag -fpie :
set(cxx\textunderscore base \textunderscore flags "-Wall -Wshadow -fpie")\\
If you have any non-standard locations where the compiler should search for Boost and/or GTest, place them in the file localpreferences.txt.\\
\subsubsection{Building}
On *nix, a build script is provided to drive cmake:\\
\lstinline!$main/generate_build.sh {Debug|Release} {g++|clang++}!\\
Alternatively, invoke cmake manually :
\lstinline!$mkdir build!\\
\lstinline!$cd build!\\
\lstinline!$cmake -G"Eclipse CDT4 - Unix Makefiles" -DCMAKE_CXX_COMPILER_ARG1=!\\
\lstinline!"{-std=c++11|-std=gnu++11}" -DCMAKE_CXX_COMPILER="{clang++|g++}" !
\lstinline!-DCMAKE_BUILD_TYPE={Release|Debug} ../main!
\lstinline!$make -j{CPU-CORES}!\\
\subsubsection{Development}
The cmake invocation has created an Eclipse CDT project for you to open the simulator code in:
\begin{itemize}
  \item Open Eclipse Luna (with CDT plugin)
  \item File, Import, General
  \item Select "Existing projects into Workspace"
  \item navigate to the root folder of the project (containing ./main, ./build)
\end{itemize}

>>>>

\section{System requirements}
\subsubsection{Compiler}
The project requires a recent version of a standards-compliant C++11 compiler and runtime. The following are compilers/runtimes used to develop the simulator (and therefore guaranteed to work)
\begin{itemize}
  \item G++ 4.8.2 + libstdc++
  \item G++ 4.9.2 + libstdc++
  \item Clang 3.4.0 + libstdc++
  \item Clang 3.5.0 + libstdc++
  \item G++ 4.9.2 on Cygwin $>=$ 1.7
\end{itemize}
Especially the C++11 threading support is essential.
\subsubsection{Dependencies}
The build system requires CMake $>=$ 2.8, the presence of standard build tools is assumed (make etc.).
The following are the functional dependencies:
\begin{itemize}
  \item Cereal (source included)
  \item GTest $>=$ 1.6
  \item Boost heap $>=$ 1.55  
\end{itemize}
Boost can be installed on *nix with the following command:\\
Debian based:\\
\lstinline!$sudo apt-get install libboost-all-dev!\\
Red-hat based:\\
\lstinline!$sudo yum install boost-devel!\\
\subsubsection{Preparing the build}
We use thread-sanitizer to verify thread-safety at runtime instrumented by the compiler (g++ and clang). This however requires building the project (and by extension any linked libraries) with position independent flags. For the simulator this is resolved by CMake, however for gtest you'll need to manually build GTest with this extra flag:\\
In file internal\textunderscore utils.cmake change compile options to append the compile flag -fpie :
set(cxx\textunderscore base \textunderscore flags "-Wall -Wshadow -fpie")\\
If you have any non-standard locations where the compiler should search for Boost and/or GTest, place them in the file localpreferences.txt.\\
SET({GTEST|BOOST}\_ROOT "your path here"). Make sure that for gtest, the listed path contains both the "include" folder and libraries in a dedicated folder "lib". For Boost the path needs to contain the "include" folder with the boost headers.
\subsubsection{Building}
On *nix, a build script is provided to drive cmake:\\
\lstinline!$main/generate_build.sh {Debug|Release} {g++|clang++}!\\
Alternatively, invoke cmake manually :
\lstinline!$mkdir build!\\
\lstinline!$cd build!\\
\lstinline!$cmake -G"Eclipse CDT4 - Unix Makefiles" -DCMAKE_CXX_COMPILER_ARG1=!\\
\lstinline!"{-std=c++11|-std=gnu++11}" -DCMAKE_CXX_COMPILER="{clang++|g++}" !
\lstinline!-DCMAKE_BUILD_TYPE={Release|Debug} ../main!
\lstinline!$make -j{CPU-CORES}!\\
\subsubsection{Development}
The cmake invocation has created an Eclipse CDT project for you to open the simulator code in:
\begin{itemize}
  \item Open Eclipse Luna (with CDT plugin)
  \item File, Import, General
  \item Select "Existing projects into Workspace"
  \item navigate to the root folder of the project (containing ./main, ./build)
\end{itemize}
